---
title: "Statistics in Data Analysis"
teaching: 0
exercises: 0
questions:
- "How can information be extracted and communicated from experimental data?"
objectives:
- "Plotting reveals information in the data."
- "Statistical significance testing compares experimental data obtained to probability distributions of data that might also be possible."
- "A probability distribution is a mathematical function that gives the probabilities of different possible outcomes for an experiment."
keypoints:
- "Plotting and significance testing describe patterns in the data and quantify effects against random variation."
source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```

## The first step in data analysis: plot the data!
A picture is worth a thousand words, and a picture of your data could reveal
important information that can guide you forward. So first, plot the data!

```{r, simulate_gen100_data}
# simulate Generation 100 resting heart rate data for the 1,567 study 
# participants
# rnorm() generates a random normal distribution from the mean and standard
# deviation values you provide
simulated_heart_rates <- rnorm(n = 1567, mean = 80, sd = 10)
hist(simulated_heart_rates, xlab = "resting heart rate")
```
## A picture is worth a thousand words

> ## Exercise 1: What does this picture tell you about resting heart rates?  
> Are there gaps in the data?  
> Are there large clusters of similar heart rate values in the data?  
> Are there apparent outliers?  
> Do the left and right tails of the data seem to mirror each other or not?
> >
> > ## Solution to Exercise 1
> > 
> > 
> > 
> {: .solution}
{: .challenge}

Now create a boxplot of the same data.

```{r, boxplot_simulated_gen100_data}
boxplot(simulated_heart_rates)
```

> ## Exercise 2: What does this boxplot tell you about resting heart rates?  
> What does the box signify?  
> What does horizontal black line dividing the box signify?  
> Are there apparent outliers?  
> How does the boxplot relate to the histogram?
> >
> > ## Solution to Exercise 2
> > 
> > 
> > 
> {: .solution}
{: .challenge}

## Plotting

## Statistical significance testing

## Probability and distributions

## The perils of p-values

## The t-test

## Confidence intervals

## Sample sizes and power curves

## Comparing standard deviations


{% include links.md %}