---
title: "Completely Randomized Designs"
teaching: 0
exercises: 0
questions:
- "What is a completely randomized design (CRD)?"
objectives:
- "CRD is the simplest experimental design."
- "In CRD, treatments are assigned randomly to experimental units."
- "CRD assumes that the experimental units are relatively homogeneous or similar."
- "CRD doesn't remove or account for systematic differences among experimental units."
keypoints:
- "CRD is a simple design that can be used when experimental are homogeneous."
source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```

## A single qualitative factor

## Analysis of variance (ANOVA)

> ## Exercise 1: Which kind of error?
> A study used to determine the effect of a drug on weight loss 
> could have the following sources of experimental error. 
> Classify the following sources as either biological, 
> systematic, or random error.  
> 1). A scale is broken and provides inconsistent readings.  
> 2). A scale is calibrated wrongly and consistently measures mice 1 gram heavier.   
> 3). A mouse has an unusually high weight compared to its experimental group (i.e., it is an outlier).  
> 4). Strong atmospheric low pressure and accompanying storms affect instrument readings, animal behavior, and indoor relative humidity.  
>
> >
> > ## Solution to Exercise 1
> > 
> > 1). random, because the scale is broken and provides any kind of random reading it comes up with (inconsistent reading)  
> > 2). systematic  
> > 3). biological  
> > 4). random or systematic; you argue which and explain why
> > 
> {: .solution}
{: .challenge}


## Equal variances and normality

```{r, random_number_generator}
exp_unit <- LETTERS
random_number <- sample(x = 100, size = 26)

# %% is the modulo operator, which returns the remainder from division
treatment <- ifelse(random_number %% 2 == 0, "chow", "high fat")
random_allocation <- data.frame(exp_unit, random_number, treatment)
random_allocation
```


```{r, random_assignment_table1}
table(random_allocation$treatment)
```



```{r, equal_numbers}

# place IDs and random numbers in data frame
equal_allocation <- data.frame(exp_unit, random_number)

# sort by random numbers (not by sample IDs)
equal_allocation <- equal_allocation[order(random_number),]

# now assign to treatment or control groups
treatment <- sort(rep(x = c("chow", "high fat"), times = 13))
equal_allocation <- cbind(equal_allocation, treatment)
row.names(equal_allocation) <- 1:26
equal_allocation
```


```{r, write_out_plan}
write.csv(equal_allocation, file = "random-assign.csv", row.names = FALSE)
```
> ## Discussion
> Why not assign treatment and control groups to samples in alphabetical order?  
> Did we really need a random number generator to obtain randomized equal groups?
>
> >
> > ## Solution 
> > 
> > 1). Scenario: One technician processed samples A through M, and a different 
> > technician processed samples N through Z.  
> > 2). Another scenario: Samples A through M were processed on a Monday, and 
> > samples N through Z on a Tuesday.  
> > 3). Yet another scenario: Samples A through M were from one strain, and 
> > samples N through Z from a different strain.    
> > 4). Yet another scenario: Samples with consecutive ids were all sibling 
> > groups. For example, samples A, B and C were all siblings, and all assigned 
> > to the same treatment.  
> > All of these cases would have introduced an effect (from the technician, the 
> > day of the week, the strain, or sibling relationships) that would confound 
> > the results and lead to misinterpretation.
> > 
> {: .solution}
{: .challenge}

## Inference

## A single quantitative factor

## Design issues



{% include links.md %}