---
title: "Randomized Complete Block Designs"
teaching: 0
exercises: 0
questions:
- "What is randomized complete block design?"
objectives:
- "A randomized complete block design randomizes treatments to experimental units within the block."
- "Blocking increases the precision of treatment comparisons."
keypoints:
- "Replication, randomization and blocking determine the validity and usefulness of an experiment."
source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```


## Design issues

> ## Exercise 1: Which kind of error?
> A study used to determine the effect of a drug on weight loss 
> could have the following sources of experimental error. 
> Classify the following sources as either biological, 
> systematic, or random error.  
> 1). A scale is broken and provides inconsistent readings.  
> 2). A scale is calibrated wrongly and consistently measures mice 1 gram heavier.   
> 3). A mouse has an unusually high weight compared to its experimental group (i.e., it is an outlier).  
> 4). Strong atmospheric low pressure and accompanying storms affect instrument readings, animal behavior, and indoor relative humidity.  
>
> >
> > ## Solution to Exercise 1
> > 
> > 1). random, because the scale is broken and provides any kind of random reading it comes up with (inconsistent reading)  
> > 2). systematic  
> > 3). biological  
> > 4). random or systematic; you argue which and explain why
> > 
> {: .solution}
{: .challenge}


## Randomized block design with a single replication


```{r, random_number_generator}
exp_unit <- LETTERS
random_number <- sample(x = 100, size = 26)

# %% is the modulo operator, which returns the remainder from division
treatment <- ifelse(random_number %% 2 == 0, "chow", "high fat")
random_allocation <- data.frame(exp_unit, random_number, treatment)
random_allocation
```

## Sizing a randomized block experiment

```{r, write_out_plan}
write.csv(equal_allocation, file = "random-assign.csv", row.names = FALSE)
```
> ## Discussion
> Why not assign treatment and control groups to samples in alphabetical order?  
> Did we really need a random number generator to obtain randomized equal groups?
>
> >
> > ## Solution 
> > 
> > 1). Scenario: One technician processed samples A through M, and a different 
> > technician processed samples N through Z.  
> > 2). Another scenario: Samples A through M were processed on a Monday, and 
> > samples N through Z on a Tuesday.  
> > 3). Yet another scenario: Samples A through M were from one strain, and 
> > samples N through Z from a different strain.    
> > 4). Yet another scenario: Samples with consecutive ids were all sibling 
> > groups. For example, samples A, B and C were all siblings, and all assigned 
> > to the same treatment.  
> > All of these cases would have introduced an effect (from the technician, the 
> > day of the week, the strain, or sibling relationships) that would confound 
> > the results and lead to misinterpretation.
> > 
> {: .solution}
{: .challenge}

## True replication

## Balanced incomplete block designs





Experimental units can be grouped, or *blocked*, to increase the precision of
treatment comparisons. Imagine that you want to evaluate the effect of different 
doses of a new drug on the proliferation of cancer cells in vitro. Divide the 
cells into four groups, each consisting of the same number of cells. Treat each 
group with a different dose of the drug for five consecutive days. For example, 
the four treatments could be:

Group 1: Control (no drug)
Group 2: Low dose (10 μM)
Group 3: Medium dose (50 μM)
Group 4: High dose (100 μM)

```{r, complete_random_block}

# create dosage levels
f <- factor(c("control", "low", "medium", "high"))
b1t <- sample(f, 4)
b2t <- sample(f, 4)
b3t <- sample(f, 4)
b4t <- sample(f, 4)
t <- c(b1t, b2t, b3t, b4t)
block <- factor(rep(c("shelf1", "shelf2", "shelf3", "shelf4"), each = 4))
cagenum <- rep(1:4, 4)
plan <- data.frame(block = block, CageNumber = cagenum, treatment = t)
plan

```
{% include links.md %}